Cosmic Microwave Background Power Spectra from Few Bit Timestreams

### 2018-09-04 digitisation.tex
Editing
Condinuted Editing.

Observations of the Cosmic Microwave Background (CMB) are of significant value to modern cosmology and particle physics. Future CMB experiments face issues in hardware design, mission planning and analysis that arise from the vast size of time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We explore the viability of reducing the TOD to few-bit numbers to address these issues. Unlike lossless compression, this technique introduces additional noise into the data. We present 1, 2 and 3 bit digitisation schemes and determine the degradation this compression causes in temperature and polarisation power spectra. We find that 3 bit digitisation has a percent-level contribution to the map noise level. We argue that such digitisation is a promising strategy for upcoming experiments.

%Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We investigate using extreme digitisation to address these issues. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by applying this process to the TOD. We find that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

\keywords{cosmic background radiation --- polarization --- data compression}
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since its discovery \citep{penzias1965}. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. Upcoming science goals include the discovery of inflationary gravitational waves and thorough studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects. Moreover, CMB experiments will measure the relativistic number of species and the neutrino mass sum \citep{s4sciencebook}.

%The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally, studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will open up access to unprecedented insight. Moreover, through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

High fidelity CMB science demands an excellent standard of data analysis. The CMB community has developed a variety of compression and computational techniques to manage the increasing influx of data, while maximising the science output \citep{tristam2007}. These include the compression of time-ordered data (TOD) into maps \citep{tegmark1997}, bandpower estimation \citep{tegmark1998} and the pseudo $C_l$ method \citep{brown2005}.

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD \citep{gaztanaga1998, maris2003}. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a review of the current compression techniques employed we are sure to lose information.

@@

This work is structured as follows. In \S\ref{sec:dig} we detail the challenges originating from handling large TOD, formulate the process of extreme digitisation and lay out the framework used to test its performance. Subsequently, we describe the power spectrum estimation employed and interpret the results obtained through this process in section \S\ref{sec:results}. We summarise our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume \citep{s4sciencebook}..

Excellent observation conditions for CMB observations can not only be found in space, but also at various locations on earth \citep{li2017, kovac2007} among which is the South Pole. Space- and Antarctica-based experiments depend on satellite transmission. The next generation of ground-based instruments will aim to collect 2 million detector years of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in said range will go unnoticed for longer.

The Planck mission has demonstrated the merits of carrying out CMB observations from space \citep{planck2018}. Upcoming missions aim to exceed the detector count of Planck by at least an order of magnitude \citep{litebird2014, pixie2011, core2018}. However, strategies to meet the telemetry specifications of each mission appear to be non-settled. It is not clear whether the data compression knowledge developed during the Planck mission will guarantee optimal performance for future satellites. Each mission will have to carefully construct a compression algorithm through a combination of lossless and lossy techniques.

%It is questionable whether their telemetry specifications will allow for transmission of the data with Planck-style compression. Methods of storing large amounts of data on upcoming satellites will likely be prohibited by financial decisions: the amount of storage space required becomes financially relevant at the scales targeted. Missions will likely be left to design their own compression algorithms which will incorporate a combination of lossless and lossy compression techniques.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive. The community depends on the excellent facilities provided by the National Energy Research Scientific Computing Center (NERSC).

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. Beyond that extreme digitisation has the potential of solving planning and analysis problems.

@@

Digitisation is a lossy compression technique. However, the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. There is usually little concern around finding an optimal set of digitisation parameters for a given input signal. It is common to have access to large numbers of bits to store information, where changes in the digitisation scheme become insignificant to the distortion induced. It is in our interest however to consider extreme, i.e. few-bit, digitisation. To do so we review the key ideas laid out by \cite{max1960} below.

Digitisation discretises an input signal by sorting it into $N$ appropriate ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. A digitisation scheme is described by the number of ranges, $N$, the endpoints of these ranges, $x_i$, and the output levels, $y_i$. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to quantify the performance of a given digitisation scheme we define the distortion as

\[ D = \left\langle  \left( s - \hat{s} \right)^2 \right\rangle \]

@@

\[ D = \sum_{i = 1}^N \int_{x_i}^{x_{i+1}} \left(x-y_i\right)^2 p(x) dx \]

Seeing as we wish to minimise the distortion we differentiate the above with respect to $x_i$ and $y_i$ and set the derivatives to zero. We obtain the two equations

\begin{equation} \label{eq:distderiv1}
\frac{\partial D}{\partial x_i} = \left(x_i-y_{i-1}\right)^2 p(x_i) - \left(x_i - y_i\right)^2 p(x_i) = 0
\end{equation}

\begin{equation} \label{eq:distderiv2}
\frac{\partial D}{\partial x_j} = -2 \int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0
\end{equation}

@@

\begin{equation} \label{eq:digitequalspacecondition}
x_i = \frac{y_i+y_{i+1}}{2}
\end{equation}

which informs us that an output level $y_i$ must lie halfway between its delimiting thresholds $x_i$ and $x_{i+1}$. We gain an additional condition from equation \ref{eq:distderiv2}

\begin{equation} \label{eq:digitareacondition}
\int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0
\end{equation}

@@

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we assume that CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise\footnote{Please see the conclusion for a discussion of the effect of more realistic noise profiles.}, i.e. $p(x) = 1/\sqrt{2\pi} e^{-x^2/2}$. Given this assumption we can solve the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one observes whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_i$'s and $y_i$'s were chosen appropriately.

We use the findings of Max to formulate the multi-level functions we use for our 1, 2 and 3 bit digitisation process. Given an input signal $s(t)$ a digitisation scheme using $N$ bits returns the output $\hat{s}_N(t)$. For 1 bit digitisation we apply the sign function

\[ \hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & \text{for } s(t) > 0\\
-1, & \text{for } s(t) \leq 0
\end{array} \right. \]

@@

\[ \hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & \text{for } s(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq s(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } 0.9816 \sigma \leq s(t) < 0\\
-1.51 \sigma, & \text{for } 0.9816 \sigma < s(t)\\
\end{array} \right. \]

where $\sigma$ is the standard deviation of the input signal. Finally the optimal 3 bit digitisation is described by the eight-level function

\[ \hat{s}_3(t) = \left\{ \begin{array}{rl}
2.152 \sigma, & \text{for } s(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq s(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 0\\
-0.756 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 0.501 \sigma\\
-1.344 \sigma, & \text{for } 1.748 \sigma \leq s(t) < 1.05 \sigma\\
-2.152 \sigma, & \text{for } 1.748 \sigma < s(t)\\
\end{array} \right. \]

@@

Completely different digitisation schemes can be thought of, which would for example reference the most drastic outlier in the dataset, or seek to place equal numbers of points into each digitisation level. However, given the assumptions made the schemes derived above are optimal. Additionally, they are simple enough to be easily implemented computationally.

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain control maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps with $\mathrm{NSIDE} = 4096$ based on the results of Planck 2015. The key cosmological parameters assumed are $\Omega_b \approx 0.049 $, $\Omega_c \approx 0.265$, $\Omega_m \approx 0.316$ and $h \approx 0.67$.

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that all pixels within the patch are hit approximately uniformly. The speed at which we sweep across the survey area is adjusted to to produce a desired number of hits per pixel (hits per pixel) in the output maps.

While performing each CES the pixels being targeted are determined. The corresponding values from the template maps are then accessed and added to realisations of the detector noise of appropriate length. We assume the detector noise to be Gaussian white noise.

@@

This process is carried out 6 times. We produce maps with approximately $800$, $8,000$, $80,000$, $1,024,000$, $10,240,000$ and $102,400,000$ hits per pixel. We assume that the detector is read out $200$ times per second and has a noise level of $500 \mu K \sqrt{s}$ for temperature and $\sqrt{2} \times 500 \mu K \sqrt{s}$ for polarisation observations. The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. To carry out this simulation we make use of parallelisation and the computing facilities provided by NERSC.

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Planck 2015\\
%\hline\\[-1.5ex]
%$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
%$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
%$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
%$H_0$ & $67.26 \pm 0.98$\\
%$n_s$ & $0.9652 \pm 0.0062$\\
%$\Omega_m$ & $0.316 \pm 0.014$\\
%$\sigma_8$ & $0.830 \pm 0.015 $\\
%$\tau$ & $0.078 \pm 0.019$\\
%$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
%\end{tabular}
%\tablecomments{ 
%Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
%} \normalsize
%\end{center}
%\end{table}

%\begin{table}[tbh] % transpose and include hits per pixel!
%\begin{center}
%\caption{\label{tab:modelparams} Assumed Survey Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Value\\
%\hline \\[-1.5ex]
%$\mathrm{NSIDE}$ &$4096$\\
%$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
%$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
%$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
%$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
%\multirow{2}{*}{$\mathrm{hits per pixel}$} & $\left(1, 10, 100 \right) \times 800,$\\
%& $\left( 1, 10, 100 \right) \times 1024000$ \\
%\end{tabular}
%\tablecomments{ 
%Parameters used in the simulated observation. The RA speed is adjusted to match the desired hits per pixel.
%} \normalsize
%\end{center}
%\end{table}

@@

We use PolSpice to compute the TT, EE and BB power spectra of the reconstructed maps I, Q, U maps. When doing so we apodise the observed skypatch using a cosine mask with $\sigma_{\mathrm{APOD}} = \sqrt{600}/2 \mathrm{deg}$ to minimise cut sky effects on the spectra.

@@

Normalisation of the obtained power spectra is necessary. The digitisation schemes introduced in section \S\ref{sec:dig} are designed to minimise the distortion of the TOD, not to preserve power at the map level. Calibration to the CMB is common for many experiments (e.g. SPT).

For the normalisation procedure we focus on the $\sim 10^8 \mathrm{hits per pixel}$ maps. We normalise the $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra from few bit TOD against their control counterparts. We consider the normalisation window cut out by the conditions

\[ l > 35 \> ; \> \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10 \]

where $C_l^{\mathrm{S}}$ is the power spectrum of the input template maps and $C_l^{\mathrm{N}}$ the detector noise level. The lower limit is placed in accordance to the patch-size surveyed, the upper limit ensures appropriate signal to noise in the normalisation window.

The normalisation constants for each digitisation scheme and channel are obtained through this way are then applied to all lower hits per pixel simulations. A sample of the obtained power spectra are shown in figure \ref{fig:powspeczoom}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
  \caption[Current ]{
   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hits per pixel}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
\label{fig:powspeczoom}
}
\end{figure}

@@

\section{Conclusions}
\label{sec:conclusions}

>>>>

Observations of the Cosmic Microwave Background (CMB) are of significant value to modern cosmology and particle physics. Future CMB experiments face issues in hardware design, mission planning and analysis that arise from the vast size of time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctica- and space-based experiments which depend on satellite links to transmit their data. We explore the viability of reducing the TOD to few-bit numbers to address these issues. Unlike lossless compression, this technique introduces additional noise into the data. We present 1, 2 and 3 bit digitisation schemes and determine the degradation this compression causes in temperature and polarisation power spectra. We find that 3 bit digitisation has a percent-level contribution to the map noise level. We argue that such digitisation is a promising strategy for upcoming experiments.

\end{abstract}

%C: maybe sub techniques:misc for methods: data analysis ? unfortunately no data compression tag
\keywords{cosmic background radiation --- polarization --- techniques: miscellaneous}
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since its discovery \citep{penzias1965}. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. Upcoming science goals include the discovery of inflationary gravitational waves, the construction of precise lensing maps and thorough studies of the Sunyaev-Zeldovich (SZ) effects. Moreover, CMB experiments will measure the relativistic number of species and the neutrino mass sum \citep{s4sciencebook}. %L: reference to science goals by satellite missions as well

High fidelity CMB science demands an excellent standard of data analysis. The CMB community has developed a variety of compression techniques and computational approaches to manage the increasing influx of data, while maximising the science output \citep{tristam2007}. These include the compression of time-ordered data (TOD) into maps \citep{tegmark1997}, bandpower estimation \citep{tegmark1998} and the pseudo $C_l$ method \citep{brown2005}.

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD \citep{gaztanaga1998, maris2003}. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a thorough review of compression techniques we are sure to lose information. %L:ref to non-planck satellites for reduced bit TOD? %C: Is there a reference detailing downsampling of SPT data?

@@

This work is structured as follows. In \S\ref{sec:dig} we detail the challenges originating from handling large TOD, introduce the process of extreme digitisation and lay out the framework used to test its performance. Subsequently, we describe the power spectrum estimation employed and interpret the obtained results in section \S\ref{sec:results}. We summarise our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume \citep{s4sciencebook}. %L: check out the particle paper where the moores law figure comes from

Excellent observation conditions for CMB observations can not only be found in space, but also at various locations on earth \citep{li2017, kovac2007} among which is the South Pole. Space- and Antarctica-based experiments depend on satellite transmission. An experiment at the South Pole will be part of the next generation of ground-based instruments, which aims to collect 2 million detector years in total. Such an experiment will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted data set. This also means that any potential faults or errors in the experiment that only become visible in said range will go unnoticed for longer. %L: reference for observing conditions in chile or a general review of sites %C: Is there a paper detailing trasmission limits for SPT and that the data is currently downsampled?

The Planck mission has demonstrated the merits of carrying out CMB observations from space \citep{planck2018}. Upcoming missions aim to exceed the detector count of Planck by at least an order of magnitude \citep{litebird2014, pixie2011, core2018}. However, strategies to meet the telemetry specifications of each mission appear to be in development. It is not clear whether the data compression knowledge developed during the Planck mission will guarantee optimal performance for future satellites. Most likely each mission will have to carefully construct a compression algorithm through a combination of lossless and lossy techniques, specific to their requirements.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive. The community depends on the facilities provided by the National Energy Research Scientific Computing Center (NERSC).

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. Beyond that extreme digitisation has the potential of easing issues in planning, analysis and hardware requirements.

@@

Digitisation is a lossy compression technique. However, the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. There is usually little concern around finding an optimal set of digitisation parameters for a given input signal. It is common to have access to large numbers of bits to store information, where changes in the digitisation scheme become insignificant to the distortion induced. It is in our interest however to consider extreme, i.e. few-bit, digitisation, where a digitisation scheme must be carefully constructed. We review the key ideas regarding construction of digitisation schemes as laid out by \cite{max1960} below.

Digitisation discretises an input signal by sorting it into $N$ appropriate ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. The set of parameters $N, x_i, y_i$ fully specify a digitisation scheme. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to quantify the performance of a given digitisation scheme we define the distortion as

\begin{equation} D = \left\langle  \left( s - \hat{s} \right)^2 \right\rangle, \end{equation}

@@

\begin{equation} D = \sum_{i = 1}^N \int_{x_i}^{x_{i+1}} \left(x-y_i\right)^2 p(x) dx. \end{equation}

Since we wish to minimise the distortion we differentiate the above with respect to $x_i$ and $y_i$ and set the derivatives to zero. We obtain the two equations

\begin{equation} \label{eq:distderiv1}
\frac{\partial D}{\partial x_i} = \left(x_i-y_{i-1}\right)^2 p(x_i) - \left(x_i - y_i\right)^2 p(x_i) = 0,
\end{equation}

\begin{equation} \label{eq:distderiv2}
\frac{\partial D}{\partial x_j} = -2 \int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0.
\end{equation}

@@

\begin{equation} \label{eq:digitequalspacecondition}
x_i = \frac{y_i+y_{i+1}}{2},
\end{equation}

which informs us that an output level $y_i$ must lie halfway between its delimiting thresholds $x_i$ and $x_{i+1}$. From equation \ref{eq:distderiv2} we gain the additional condition

\begin{equation} \label{eq:digitareacondition}
\int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0.
\end{equation}

@@

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we assume that CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise with standard deviation $\sigma$, i.e. $p(x) = (1/\sqrt{2\pi\sigma^2}) e^{-x^2/2\sigma^2}$. Given this assumption we can solve the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one checks whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_i$'s and $y_i$'s were chosen appropriately.

We use the findings of \cite{max1960} to formulate the multi-level functions we use for our 1, 2 and 3 bit digitisation process. Given an input signal $s(t)$ a digitisation scheme using $N$ bits returns the output $\hat{s}_N(t)$. For 1 bit digitisation we apply the sign function

\begin{equation} \hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & \text{for } s(t) > 0\\
-1, & \text{for } s(t) \leq 0
\end{array} \right. \end{equation}

@@

\begin{equation} \hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & \text{for } s(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq s(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } 0.9816 \sigma \leq s(t) < 0\\
-1.51 \sigma, & \text{for } 0.9816 \sigma < s(t)\\
\end{array} \right. . \end{equation}
%C: should i round to equal significant figures here? Unfortunately these are the exact values that i used...

Finally the optimal 3 bit digitisation is described by the eight-level function

\begin{equation} \hat{s}_3(t) = \left\{ \begin{array}{rl}
2.152 \sigma, & \text{for } s(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq s(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 0\\
-0.756 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 0.501 \sigma\\
-1.344 \sigma, & \text{for } 1.748 \sigma \leq s(t) < 1.05 \sigma\\
-2.152 \sigma, & \text{for } 1.748 \sigma < s(t)\\
\end{array} \right. . \end{equation}

@@

Completely different digitisation schemes can be thought of, which would for example reference the most drastic outlier in the data set, or seek to place equal numbers of points into each digitisation level. However, given the assumptions made the schemes derived above are optimal. Additionally, they are simple enough to be easily implemented computationally.

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain control maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.
%C: Do I need to mention that we assume to have an infinitely fast spinning plate? %L: Check whether Python uses 32 or 64 bit under on NERSC for floats

To create the template maps we use the Hierarchical Equal Area isoLatitude Pixelization of a sphere \citep[HEALPix;][]{healpix}\footnote{Available at \url{http://healpix.sourceforge.net/}} framework. We generate a realisation of I, Q and U maps with $\mathrm{NSIDE} = 4096$ based on the results of the Planck collaboration with key cosmological parameters of $\Omega_b \approx 0.049 $, $\Omega_c \approx 0.265$, $\Omega_m \approx 0.316$ and $h \approx 0.67$ \citep{planck2016}.

%C: Do I need to comment on the fact that this reflects no particular frequency channel, but the results come from a combination of all observed channels?

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that each pixel is sampled approximately uniformly. The speed at which we sweep across the survey area is adjusted to produce a desired number of hits per pixel in the output maps.

While performing each CES the pixels being targeted are determined. The corresponding values from the template maps are then accessed and added to realisations of the detector noise of appropriate length to generate TOD streams. We assume the detector noise to be Gaussian white noise. For temperature observations we assume the detector noise level to be $500 \mathrm{\mu K \sqrt{s}}$ and for polarisation observations $\sqrt{2} \hspace{0.7ex} 500 \mathrm{\mu K \sqrt{s}}$, i.e. the corresponding photon noise limit.

@@

% LINK TO GITHUB WITH CODE HERE

This process is carried out 6 times. We produce maps where pixels in the survey patch are hit approximately uniformly $800$, $8,000$, $80,000$, $1,024,000$, $10,240,000$ and $102,400,000$ times. We assume that the detector is read out $200$ times per second. %The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. This calculation was made possible through use of the NERSC facilities.

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Planck 2015\\
%\hline\\begin{equation}-1.5ex]
%$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
%$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
%$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
%$H_0$ & $67.26 \pm 0.98$\\
%$n_s$ & $0.9652 \pm 0.0062$\\
%$\Omega_m$ & $0.316 \pm 0.014$\\
%$\sigma_8$ & $0.830 \pm 0.015 $\\
%$\tau$ & $0.078 \pm 0.019$\\
%$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
%\end{tabular}
%\tablecomments{ 
%Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
%} \normalsize
%\end{center}
%\end{table}

%\begin{table}[tbh] % transpose and include hits per pixel!
%\begin{center}
%\caption{\label{tab:modelparams} Assumed Survey Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Value\\
%\hline \\begin{equation}-1.5ex]
%$\mathrm{NSIDE}$ &$4096$\\
%$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
%$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
%$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
%$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
%\multirow{2}{*}{$\mathrm{hits per pixel}$} & $\left(1, 10, 100 \right) \times 800,$\\
%& $\left( 1, 10, 100 \right) \times 1024000$ \\
%\end{tabular}
%\tablecomments{ 
%Parameters used in the simulated observation. The RA speed is adjusted to match the desired hits per pixel.
%} \normalsize
%\end{center}
%\end{table}

@@

We use the Spatially Inhomogeneous Correlation Estimator for Temperature and Polarisation \citep[PolSpice;][]{polspice} to compute the $\mathrm{TT}, \mathrm{EE}$ and $\mathrm{BB}$ power spectra of the reconstructed maps I, Q, U maps. To minimise cut sky effects we apodise the observed path using a cosine mask with $\sigma_{\mathrm{APOD}} = \sqrt{600}/2 \hspace{0.7ex} \mathrm{deg}$. %C: Should I also mention the multiplication with a cosine mask that I supply in addition to setting this option on polspice? This will likely be reworked when using Healpix to get the power spectra though.

%%%%%%%%%%%
% INCLUDE SECTION OF ALM CUTS (maybe a plot of the cuts too)
%%%%%%%%%%%

@@

%%%%%%%%%%%
% THIS MIGHT CHANGE WHEN WORKING OUT THEORETICAL NORMALISATION
%%%%%%%%%%%

Normalisation of the obtained power spectra is necessary. The digitisation schemes introduced in section \S\ref{sec:dig} are designed to minimise the distortion of the TOD, not to preserve power at the map level. Normalising obtained spectra against previous experiments is a common calibration technique in many CMB experiments (e.g. SPT). %C: Do you know a good reference for this?

For the normalisation procedure we focus on the $\sim 10^8$ hits per pixel maps. We normalise the $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra from few bit TOD against their control counterparts. We consider the normalisation window cut out by the conditions

\begin{equation} l > 35 \> , \> \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10. \end{equation}

where $C_l^{\mathrm{S}}$ is the power spectrum of the input template maps and $C_l^{\mathrm{N}}$ the detector noise level. The lower limit is placed in accordance to the patch-size surveyed while the upper limit ensures appropriate signal to noise in the normalisation window.

The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hits per pixel simulations. A sample of the obtained power spectra is shown in figure X.

%%%%%%%%%%%
%SAMPLE POWER SPECTRA WITH THEORY SPECTRA
% want to show that we recover the PS, including theory PS in background
%%%%%%%%%%%

%\begin{figure}[htb]\centering
%\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
%\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
%\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
%  \caption[Current ]{
%   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hits per pixel}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
%\label{fig:powspeczoom}
%}
%\end{figure}

@@

\section{Conclusion}
\label{sec:conclusions}
